Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] minor typo
[ paper ] more fixes

As usual, definitional equality of types impacts typing of terms via
the conversion rule:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : S \\ 0\Gamma \vdash S \equiv T~\istype}
  {\Gamma \vdash M : T}
\end{displaymath}
Like type formation, definitional equality of types always takes place
in $0$d contexts. We will describe the definitional equality rules for
terms of each type as we introduce them. In QTT, it is possible for
the definitional equality of terms to differ between the $\sigma = 0$
and $\sigma = 1$ fragments, as we will see below.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$, where $x$ does not appear in
$T$. $\Pi$-types support the usual $\beta\eta$ definitional
equalities.

@@

The equality rules for $\Rtype$ state that the two operations are
mutually inverse: $\rIntro(\rElim(M)) \equiv M$, in both fragments,
and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment.

@@

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{AroraBarak09}, where the correctness of
implementation is probabilistic. With a dependently-typed host
language, adding probabilistic choice as an effect, and capturing the
semantic constraint of BPP is straightforward.

>>>>

As usual, definitional equality of types impacts typing of terms via
the conversion rule:
\begin{displaymath}
  \inferrule*
  {\Gamma \vdash M : S \\ 0\Gamma \vdash S \equiv T~\istype}
  {\Gamma \vdash M : T}
\end{displaymath}
Like type formation, definitional equality of types always takes place
in $0$-d contexts. We will describe the definitional equality rules for
terms of each type as we introduce them. In QTT, it is possible for
the definitional equality of terms to differ between the $\sigma = 0$
and $\sigma = 1$ fragments, as we will see below.

@@

  \inferrule*
  {\Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\rho: S) \to T \\
    \Gamma_2 \vdash N \stackrel{\sigma'}: S \\
    0\Gamma_1 = 0\Gamma_2 \\
    \sigma' = 0 \Leftrightarrow (\rho = 0 \lor \sigma = 0)}
  {\Gamma_1 + \rho\Gamma_2 \vdash M\,N \stackrel\sigma: T[N/x]}
\end{mathpar}
The side conditions on the elimination rule state that (i) both
$\Gamma_1$ and $\Gamma_2$ erase to the same context, so their sum is
defined; and (ii) the argument $N$ is erased (i.e., $\sigma' = 0$) iff
either the function does not use its argument, or we are in the
$\sigma = 0$ fragment and everything is being erased. In the
following, when we write $S \to T$ for a non-dependent function type,
we mean that the argument is to be used linearly:
$(x \stackrel1: S) \to T$, where $x$ does not appear in
$T$. $\Pi$-types support the usual $\beta\eta$ definitional
equalities in both the $\sigma = 0$ and $\sigma = 1$ fragments.

@@

The equality rules for $\Rtype$ state that the two operations are
mutually inverse: $\rIntro(\rElim(M)) \equiv M$, in both fragments,
and $\rElim(\rIntro(M)) \equiv M$ in the $\sigma = 0$ fragment. By
congruence, the $\sigma = 1$ fragment's definitional equality affects
the definitional equality of the $\sigma = 0$ fragment via the
$\rIntro(-)$ constructor.

@@

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{AroraBarak09}, where the correctness of
implementation is probabilistic. With a dependently-typed host
language, adding probabilistic choice as an effect and capturing the
semantic constraint of BPP is straightforward.
